Array

### 2013-11-19 PAPERS/Report/AdvancedTypes/AdvancedTypes.tex
Fixing some grammar problems and adding a presentation produced last week.

@@

Two approaches are possibles to support arrays. An approach is importing
reusable libraries of AtelierB that are reported at official document of
ClearSy\footnote{\url{http://www.atelierb.eu/ressources/DOC/english/reusable-components-reference.pdf}}.
This approach is supported, but the generated code is more complex and it needs
more intermediate steps to manipulate the data of imported components. Not only,
this approach allows dynamic allocation for arrays, but it is extremely not
recommended to safety-critical systems. A second approach enjoys the direct 
correspondence between arrays native B0 and LLVM-IR and this approach is presented here.

A concrete array is a total function, where the domain indexes the range 
elements. The domain is composed by Cartesian product of sequence of concrete 
sets. The node Data array in AST can represent a declaration of array or a 
selection of element. The first element ``dec'' declares explicitly an array or does a 
reference to a variable array. The second element ``type'' defines a sequence of 
array types, where the last element defines the range of function and the others 
elements defines the domain. The third element ``indexes'' is optional, and it 
represents the selection of an element in array.

@@

The B0 language supports attributions between arrays, even though the attribution
must have the same domain type and range type. 
%TODO to add this restriction in become equals.

@@

The arrays in B0 are represented by functions where an index pointer to an 
element. By default, the element of index not initialized has the zero value.

@@

The first parameter defines the type and name of variable, and the others are
indexes to select the element in the data type. The second parameter (i64 0)
selects the first element in the pointer (* @Array\_\_arr). The last element,
(i64 \%3) selects the value in position obtained from \%3 of the array of 32-bit
integer values. When the array is multidimensional, news index must be added
separated by comma to get the specific element.

>>>>

\section{The LLVM grammar with support to array multidimensional}

\begin{figure}
  \begin{center}
    \begin{tabular}{rcl}
      \nt{module} & ::= & \ListOf{\nt{item}} \\
      \nt{item} & ::= & \nt{const\_decl} \alt \nt{function\_decl} \alt \nt{type\_def}
      \alt \nt{const\_def} \alt \nt{var\_def} \alt \nt{function\_def} \\
      \nt{const\_decl} & ::= & \nt{name} \llvm{=} \llvm{external} \llvm{constant} \nt{type} \\
      \nt{type\_def} & ::= & \nt{name} \llvm{=} \llvm{type} \nt{type} \\
      \nt{type} & ::= & \llvm{void} \alt \nt{itype} \alt \llvm{\{} \ListOf{\nt{type}} \llvm{\}} \alt \nt{type}\llvm{*} \\
      \nt{const\_def} & ::= & \nt{name} \llvm{=} \llvm{constant} \nt{type} \nt{iliteral} \\
      \nt{var\_def} & ::= & \nt{name} \llvm{=} \llvm{common} \llvm{global} \nt{type} \llvm{zeroinitializer} \\
      \nt{function\_decl} & ::= & \llvm{declare} \nt{type} \nt{name} \llvm{(} \ListOf{\nt{type}} \llvm{)}\\
      \nt{function\_def} & ::= & \llvm{define} \nt{type} \nt{name} \llvm{(} \ListOf{\nt{param}} \llvm{)} \llvm{\{} \ListOf{\nt{block}} \llvm{\}} \\
      \nt{param} & ::= & \nt{type} \nt{name} \\
      \nt{block} & ::= & \nt{lbl} \llvm{:} \ListOf{\nt{inst}} \\
      \nt{inst} & ::=  & \nt{name} \llvm{=} \llvm{alloca} \nt{type} \\
      & \alt & \nt{name} \llvm{=} \lalt \llvm{add} \alt \llvm{sub} \alt \llvm{mul} \alt \llvm{sdiv} \alt \llvm{srem} \ralt \nt{itype} \nt{exp} \llvm{,} \nt{exp} \\
      & \alt & \nt{name} \llvm{=} \llvm{icmp} \lalt \llvm{eq} \alt \llvm{ne} \alt \llvm{sgt} \alt \llvm{sge} \alt \llvm{slt} \alt \llvm{sle} \ralt \llvm{i1} \nt{exp} \llvm{,} \nt{exp}\\
      & \alt & \nt{name} \llvm{=} \llvm{call} \nt{type} \llvm{(} \ListOf{\nt{arg}} \llvm{)} \\
      & \alt & \nt{name} \llvm{=} \llvm{getelementptr} \nt{type} \llvm{*} \nt{exp}\llvm{,} \nt{index}\llvm{,} \nt{index} \\
      & \alt & \nt{name} \llvm{=} \llvm{load} \nt{type} \nt{exp} \\
      & \alt & \llvm{store} \nt{type} \nt{exp}, \nt{type} \llvm{*} \nt{exp} \\
      & \alt & \llvm{br} \llvm{i1} \nt{exp} \llvm{,} \llvm{label} \nt{lbl} \llvm{,} \llvm{label} \nt{lbl} \\
      & \alt & \llvm{br} \llvm{label} \nt{lbl} \\
%       & \alt & \llvm{switch} \nt{type} \nt{exp} \llvm{,} \llvm{branch} \nt{lbl} \llvm{[} \ListOf{\nt{branch}} \llvm{]} \\
      & \alt & \llvm{ret} \lalt \nt{type} \nt{exp} \alt \llvm{void} \ralt \\
      \nt{exp} & ::= & \nt{name} \alt \nt{iliteral} \alt \llvm{getelementptr} \llvm{(} \nt{type} \nt{exp} \llvm{,} \nt{index} \llvm{,} \nt{index} \llvm{)} \\
      \nt{index} & ::= & \nt{itype} \nt{iliteral} \\
      \nt{branch} & ::= & \nt{iliteral} \nt{iliteral} \nt{lbl} \\
      \nt{arg} & ::= & \nt{type} \nt{exp}
    \end{tabular}
  \end{center}
  \caption{Grammar of the target LLVM subset: \nt{itype}, \nt{iliteral}, \nt{lbl}
    and \nt{name} correspond respectively to integer types, integer literals,
    labels and names. Choices are separated by \alt and optionally delimited by
    \lalt and \ralt.  The \ListOf{} superscript denotes a comma-separated list of
    elements of the annotated entity.}
  \label{fig:llvm-grammar}
\end{figure}

@@

Two approaches are possible to support arrays. An approach is importing
reusable libraries of AtelierB that are reported in the official document of
ClearSy\footnote{\url{http://www.atelierb.eu/ressources/DOC/english/reusable-components-reference.pdf}}.
This approach is supported, but the generated code is more complex and it needs
more intermediate steps to manipulate the data of imported components. Moreover,
this approach allows dynamic allocation for arrays, but it is strongly not
recommended to safety-critical systems. A second approach enjoys the direct 
correspondence between B0 native arrays and LLVM-IR and this approach is presented here.

A concrete array is a total function where the domain indexes the range 
elements. The domain is composed of Cartesian product of sequence of concrete 
sets. The node Data array in AST can represent a declaration of array or a 
selection of element. The first element ``dec'' declares explicitly an array or 
references a variable array. The second element ``type'' defines a sequence of 
array types, where the last element defines the range of function and the other 
elements define the domain. The third element ``indexes'' is optional, and it 
represents the selection of an element in array, or the same type structure 
when it is a structure.

@@

\paragraph{Attributions between array and structure}

The B0 language supports attribution directly between arrays or structure, even though the attribution
must have the same domain type and range type. \VGM{Project Decision: How translate a direct 
attribution between advanced types, for example, arrayA := arrayB ? The Commenc 
uses a function called ``memmove''. }

%TODO to add this restriction in become equals.

@@

The arrays in B0 are represented by functions where an index points to an 
element. By default, the element of index not initialized has a zero value.

@@

The first parameter defines the type and name of the variable, and the others are
indexes to select the element in the data type. The second parameter (i64 0)
selects the first element in the pointer (* @Array\_\_arr). The last element,
(i64 \%3) selects the value in position obtained from \%3 of the array of 32-bit
integer values. When the array is multidimensional, the news index must be added
separated by comma to get the specific element.
